Supermartingales and Termination Analysis of Statistical PCF

### 2021-01-24 LICS21/SPCF.tex
Use \defn to make key definienda stand out - easier for reviewers to find definitions

\subsection{Trace semantics}
\label{sec:sampling semantics}
This version of the reduction relation allows $\tsample$ to reduce to any number in $[0,1]$. 
To more precisely specify the probabilities, an additional argument is needed to determine the outcome of random samples. Let $ I := [0,1] \subset \mathbb{R} $, and define $\entrosp := I^{\mathbb{N}}$, with the Borel $\sigma$-algebra $\Sigma_\entrosp$. Equivalently, a basis of measurable sets is $\prod_{i=0}^\infty X_i$ where $X_i$ are all Borel and all but finitely many are $I$, and the probability measure $\mu_{\entrosp}$ is given by $\mu_{\entrosp} (\prod_{i=0}^\infty X_i) := \prod_{i=0}^\infty \Leb(X_i)$, \changed[lo]{writing $\Leb$ for the Lebesgue measure.}
\changed[lo]{The maps $\pi_h:\entrosp \to I$ (projecting to the first element) and $\pi_t:\entrosp \to \entrosp$ (popping the first element) are then measurable.}
Following \cite{DBLP:conf/esop/CulpepperC17}, we call the probability space $(\entrosp, \Sigma_\entrosp, \mu_{\entrosp})$ the \emph{entropy space}, and elements of $\entrosp$ \emph{traces}.

Define a \emph{skeleton} to be a term but, instead of having real constants $\underline r$, it has a placeholder $\skeletonPlaceholder$, so that each term $M$ has a skeleton $\mathit{Sk}(M)$, and each skeleton $S$ can be converted to a term $S[r]$ given a vector $r$ of $n$ real numbers to substitute in, where $n$ is the number of occurrences of $\skeletonPlaceholder$ in $S$. 
Following \cite{DBLP:conf/icfp/BorgstromLGS16,DBLP:conf/lics/StatonYWHK16,DBLP:journals/pacmpl/EhrhardPT18}, the $\sigma$-algebra on $\Lambda$ is defined by identifying $\Lambda$ with $\bigcup_{m\geq 0} \big(\mathsf{Sk}_m \times \Real^m \big)$, where $\mathsf{Sk}_m$ is the set of skeletons containing $m$ occurrences of $\skeletonPlaceholder$.
Thus identified, we give $\Lambda$ the countable disjoint union topology of the product topology of the discrete topology on $\mathsf{Sk}_m$ and the standard topology on $\Real^m$, and take the corresponding Borel $\sigma$-algebra.
Note that the connected components of $\Lambda$ have the form $\{S\} \times \Real^m$, with $S$ ranging over $\mathsf{Sk}_m$, and $m$ over $\mathbb{N}$.

@@

The reduction function is measurable, and the set of values is measurable, therefore the set of $s$ such that $M$ terminates at $s$ within $n$ steps is measurable for any $n$, therefore $\{s \mid \hbox{$M$ terminates at $s$}\}$ is measurable \cite{DBLP:conf/icfp/BorgstromLGS16,MakOPW21}. 
We say that a term $M$ is \emph{almost surely terminating} (AST) just if $\mu_{\entrosp}(\{s \mid \hbox{$M$ terminates at $s$}\}) = 1$.

\iffalse
\lo{Alternatively, define the \emph{runtime of $M$} to be the random variable 
\[
T_M(s) := 
\begin{cases}
\min \set{n \mid \pi_0(\red^n(M, s)) \textrm{ is a value}} & \hbox{if $\red^\infty(M,s)$ is defined}\\
\infty & \hbox{otherwise}
\end{cases}
\]
Equivalently, we say that $M$ is \emph{almost surely terminating} (AST) if $T_M < \infty$ a.s.; 
and $M$ is \emph{positively almost surely terminating} (PAST) if $\expect{T_M} < \infty$.}
\fi
\begin{example} The term 
\[
(\tY \lambda f n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)} ) \, \underline{0},
\] 
which generates a geometric distribution, terminates on the set $\entrosp \setminus [0.5,1]^\mathbb N$, which has measure 1, therefore it terminates almost surely; whereas 
\(
\tif{\tsample - 0.5 < 0}{\underline 0}{(\tY \lambda x. x) \, \underline 0}
\), 
which terminates on the set $\pi_h^{-1}[[0,0.5]]$, has probability 0.5 of failing to terminate.
\end{example}
%\paragraph{}
%The requirement that $M$ be closed is not important for the proofs of the theorems, but is simply needed to ensure that $M$ reaches a value, rather than a general normal form like $x \, y$. \lo{This can be omitted.}

>>>>

\subsection{Trace semantics}
\label{sec:sampling semantics}
This version of the reduction relation allows $\tsample$ to reduce to any number in $[0,1]$. 
To more precisely specify the probabilities, an additional argument is needed to determine the outcome of random samples. Let $ I := [0,1] \subset \mathbb{R} $, and define $\entrosp := I^{\mathbb{N}}$, with the Borel $\sigma$-algebra $\Sigma_\entrosp$. Equivalently, a basis of measurable sets is $\prod_{i=0}^\infty X_i$ where $X_i$ are all Borel and all but finitely many are $I$, and the probability measure $\mu_{\entrosp}$ is given by $\mu_{\entrosp} (\prod_{i=0}^\infty X_i) := \prod_{i=0}^\infty \Leb(X_i)$, \changed[lo]{writing $\Leb$ for the Lebesgue measure.}
\changed[lo]{The maps $\pi_h:\entrosp \to I$ (projecting to the first element) and $\pi_t:\entrosp \to \entrosp$ (popping the first element) are then measurable.}
Following \cite{DBLP:conf/esop/CulpepperC17}, we call the probability space $(\entrosp, \Sigma_\entrosp, \mu_{\entrosp})$ the \emph{entropy space}, and elements of $\entrosp$ \defn{traces}.

Define a \defn{skeleton} to be a term but, instead of having real constants $\underline r$, it has a placeholder $\skeletonPlaceholder$, so that each term $M$ has a skeleton $\mathit{Sk}(M)$, and each skeleton $S$ can be converted to a term $S[r]$ given a vector $r$ of $n$ real numbers to substitute in, where $n$ is the number of occurrences of $\skeletonPlaceholder$ in $S$. 
Following \cite{DBLP:conf/icfp/BorgstromLGS16,DBLP:conf/lics/StatonYWHK16,DBLP:journals/pacmpl/EhrhardPT18}, the $\sigma$-algebra on $\Lambda$ is defined by identifying $\Lambda$ with $\bigcup_{m\geq 0} \big(\mathsf{Sk}_m \times \Real^m \big)$, where $\mathsf{Sk}_m$ is the set of skeletons containing $m$ occurrences of $\skeletonPlaceholder$.
Thus identified, we give $\Lambda$ the countable disjoint union topology of the product topology of the discrete topology on $\mathsf{Sk}_m$ and the standard topology on $\Real^m$, and take the corresponding Borel $\sigma$-algebra.
Note that the connected components of $\Lambda$ have the form $\{S\} \times \Real^m$, with $S$ ranging over $\mathsf{Sk}_m$, and $m$ over $\mathbb{N}$.

@@

The reduction function is measurable, and the set of values is measurable, therefore the set of $s$ such that $M$ terminates at $s$ within $n$ steps is measurable for any $n$, therefore $\{s \mid \hbox{$M$ terminates at $s$}\}$ is measurable \cite{DBLP:conf/icfp/BorgstromLGS16,MakOPW21}. 
We say that a term $M$ is \defn{almost surely terminating} (\defn{AST}) just if $\mu_{\entrosp}(\{s \mid \hbox{$M$ terminates at $s$}\}) = 1$.

\iffalse
\lo{Alternatively, define the \emph{runtime of $M$} to be the random variable 
\[
T_M(s) := 
\begin{cases}
\min \set{n \mid \pi_0(\red^n(M, s)) \textrm{ is a value}} & \hbox{if $\red^\infty(M,s)$ is defined}\\
\infty & \hbox{otherwise}
\end{cases}
\]
Equivalently, we say that $M$ is \emph{almost surely terminating} (AST) if $T_M < \infty$ a.s.; 
and $M$ is \emph{positively almost surely terminating} (PAST) if $\expect{T_M} < \infty$.}
\fi
\begin{example}[Geometric distribution] The term 
\[
(\tY \lambda f n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)} ) \, \underline{0},
\] 
which generates a geometric distribution, terminates on the set $\entrosp \setminus [0.5,1]^\mathbb N$, which has measure 1, therefore it terminates almost surely; whereas 
\(
\tif{\tsample - 0.5 < 0}{\underline 0}{(\tY \lambda x. x) \, \underline 0}
\), 
which terminates on the set $\pi_h^{-1}[[0,0.5]]$, has probability 0.5 of failing to terminate.
\end{example}
%\paragraph{}
%The requirement that $M$ be closed is not important for the proofs of the theorems, but is simply needed to ensure that $M$ reaches a value, rather than a general normal form like $x \, y$. \lo{This can be omitted.}

### 2021-01-24 LICS21/antitone.tex
Various minors
Use \defn to make key definienda stand out - easier for reviewers to find definitions

\begin{definition}
The definition of \emph{antitone ranking function} $f$ for $M \in \Lambda^0$ is the same as that of ranking function except that in the case of $\tY$-redex, 
we require the existence of an antitone (meaning: $r < r'$ implies $\epsilon(r) \geq \epsilon(r')$) function $\epsilon : \nnReal \to \pReal$ such that the ranking function $f :\mathit{Rch}(M) \to \nnReal$ satisfies
\[f(E[R']) \leq f(E[R]) - \epsilon(f(E[R])) \] 
where $R \to R'$ is the $\tY$-redex rule.
Any closed term for which an antitone ranking function exists is called \emph{antitone rankable}.
\iffalse
Given a ranking function $f$ on $M \in \Lambda^0$, we say that $f$ is \emph{antitone} if there exists an antitone function\footnote{i.e.~$r < r'$ implies $\epsilon(r) \geq \epsilon(r')$} $\epsilon : \nnReal \to \pReal$ such that %the ranking function $f$ satisfies
\[
f(E[R']) \leq f(E[R]) - \epsilon(f(E[R])) 
\]
for all $E[R'] \in \mathit{Rch}(M)$ where $R \to R'$ is the $\tY$-redex rule.
\fi
\end{definition}

@@

\begin{definition}
Given a probability space $(\Omega, \calF, \mathbb{P})$, and a supermartingale $(Y_n)_{n \geq 0}$ and a stopping time $T$ adapted to filtration $(\calF_n)_{n \geq 0}$,
we say that $(Y_n)_{n \geq 0}$ is an \emph{antitone strict supermartingale w.r.t.~$T$} if for all $n \geq 0$, we have $Y_n \geq 0$, and there exists an antitone function $\epsilon : \nnReal \to \pReal$ satisfying
\(
\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon (Y_n) \cdot {\bf 1}_{\set{T > n}}
\).
%Any closed term for which an antitone ranking function exists is called \emph{antitone rankable}.
\end{definition}

@@

\begin{definition}
Define an \emph{antitone sparse ranking function} on a closed term $M$ to be a partial function $F : Rch(M) \rightharpoonup \mathbb{R}$ such that for some antitone function $\epsilon : \mathbb{R}_{\geq 0} \to \mathbb{R}_{>0}$:
\begin{inparaenum}[(i)]
\item $f(N) \geq 0$ for all $N \in \dom(f)$,
\item $M \in \dom(f)$,
\item for any $N \in \dom(f)$, evaluation of $N$ will eventually reach some $O$ which is either a value or in $\dom(f)$, and $f(N) \geq \mathbb E [f(O) + \epsilon(f(O)) \times \text{the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of $\dom(f)$).
\end{inparaenum}
\end{definition}

@@

\begin{example}[Non-affine recursion]
\label{ex:non-affine recursion easy}
\[
(\tY \lambda f x. x \oplus_{2/3} f (f (x + 1)))\ \underline 1
\]
\end{example}

\begin{example}[More complex non-affine recursion]
\label{ex:non-affine continuous}
\begin{align*}
& (\tY\, \lambda f x.\, (\lambda e. X[e]) \tsample)\ \underline 0 \\
& X[e] = \tif{e \leq p - x^{-2}}{x}{f^2(x+1) \oplus_e f(x+1)}
\end{align*}
\end{example}

>>>>

\begin{definition}
The definition of \defn{antitone ranking function} $f$ for $M \in \Lambda^0$ is the same as that of ranking function except that in the case of $\tY$-redex, 
we require the existence of an antitone (meaning: $r < r'$ implies $\epsilon(r) \geq \epsilon(r')$) function $\epsilon : \nnReal \to \pReal$ such that the ranking function $f :\mathit{Rch}(M) \to \nnReal$ satisfies
\[f(E[R']) \leq f(E[R]) - \epsilon(f(E[R])) \] 
where $R \to R'$ is the $\tY$-redex rule.
Any closed term for which an antitone ranking function exists is called \defn{antitone rankable}.
\iffalse
Given a ranking function $f$ on $M \in \Lambda^0$, we say that $f$ is \emph{antitone} if there exists an antitone function\footnote{i.e.~$r < r'$ implies $\epsilon(r) \geq \epsilon(r')$} $\epsilon : \nnReal \to \pReal$ such that %the ranking function $f$ satisfies
\[
f(E[R']) \leq f(E[R]) - \epsilon(f(E[R])) 
\]
for all $E[R'] \in \mathit{Rch}(M)$ where $R \to R'$ is the $\tY$-redex rule.
\fi
\end{definition}

@@

\begin{definition}
Given a probability space $(\Omega, \calF, \mathbb{P})$, and a supermartingale $(Y_n)_{n \geq 0}$ and a stopping time $T$ adapted to filtration $(\calF_n)_{n \geq 0}$,
we say that $(Y_n)_{n \geq 0}$ is an \defn{antitone strict supermartingale w.r.t.~$T$} if for all $n \geq 0$, we have $Y_n \geq 0$, and there exists an antitone function $\epsilon : \nnReal \to \pReal$ satisfying
\(
\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon (Y_n) \cdot {\bf 1}_{\set{T > n}}
\).
%Any closed term for which an antitone ranking function exists is called \emph{antitone rankable}.
\end{definition}

@@

\begin{definition}
Define an \defn{antitone sparse ranking function} on a closed term $M$ to be a partial function $F : Rch(M) \rightharpoonup \mathbb{R}$ such that for some antitone function $\epsilon : \mathbb{R}_{\geq 0} \to \mathbb{R}_{>0}$:
\begin{inparaenum}[(i)]
\item $f(N) \geq 0$ for all $N \in \dom(f)$,
\item $M \in \dom(f)$,
\item for any $N \in \dom(f)$, evaluation of $N$ will eventually reach some $O$ which is either a value or in $\dom(f)$, and $f(N) \geq \mathbb E [f(O) + \epsilon(f(O)) \times \text{the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of $\dom(f)$).
\end{inparaenum}
\end{definition}

@@

\begin{example}[Non-affine recursion]
\label{ex:non-affine recursion easy}
\[
(\tY \lambda f x. x \oplus_{2/3} f (f (x + 1)))\, \underline 1
\]
\end{example}

\begin{example}[More complex non-affine recursion]
\label{ex:non-affine continuous}
\begin{align*}
& (\tY\, \lambda f x.\, (\lambda e. X[e]) \, \tsample)\, \underline 0 \\
& X[e] := \tif{e \leq p - x^{-2}}{x}{f^2(x+1) \oplus_e f(x+1)}
\end{align*}
\end{example}

### 2021-01-24 LICS21/confluent.tex
Use \defn to make key definienda stand out - easier for reviewers to find definitions

A \emph{position} is a finite sequence of steps into a term, defined inductively as
\begin{align*}
\alpha ::= & \; {\cdot} \mid \lambda ; \alpha \mid @_1 ; \alpha \mid @_2 ; \alpha \mid \underline f_i ; \alpha \\
& {} \mid \tY ; \alpha \mid \textsf{if}_1 ; \alpha \mid \textsf{if}_2 ; \alpha \mid \textsf{if}_3 ; \alpha.
\end{align*}
The \emph{subterm of $M$ at $\alpha$}, denoted $M \mid \alpha$, is defined as
\begin{align*}
M \mid \cdot & = M \\
\lambda x. M \mid \lambda ; \alpha & = M \mid \alpha \\
M_1 M_2 \mid @_i ; \alpha & = M_i \mid \alpha \quad \text{for } i = 1,2 \\
\underline f(M_1,\dots,M_n) \mid \underline f_i ; \alpha & = M_i \mid \alpha \quad \text{for }i \leq n \\
\tY M \mid \tY ; \alpha & = M \mid \alpha \\
\tif{M_1 < 0}{M_2}{M_3} \mid \textsf{if}_i ; \alpha & = M_i \mid \alpha \quad \text{for } i = 1,2,3
\end{align*}
so that every subterm is located at a unique position, but not every position corresponds to a subterm (e.g. $x \, y \mid \lambda$ is undefined). 
A position such that $M\mid \alpha$ does exist is said to \emph{occur} in $M$. 
%\changed[lo]{
\emph{Substitution} of $N$ at position $\alpha$ in $M$, written $M[N/\alpha]$, is defined similarly.
For example, let 
\(
M = \lambda x \, y. y \, (\tif{x < 0}{y \, (\underline f (x))}{\underline 3})
\)
and
\(
\alpha =\lambda ; \lambda ; @_2 ; \textsf{if}_2 ; @_2
\)
then 
\(
M[\tsample / \alpha] = \lambda x \, y. y \, (\tif{x < 0}{y \, \tsample}{\underline 3}).
\)
%}

@@

With this notation, a more general reduction relation $\to$ can be defined.
\begin{definition}
\label{def:more general red}
The binary relation $\to$ is defined by the following rules, each is conditional on a redex occurring at position $\alpha$ in the term $M$:
\begin{align*}
  \text{if } M \mid \alpha = (\lambda x.N) V,\ & M \to M[N[V/x]/\alpha] \\
  \text{if } M \mid \alpha = \underline f (\underline r_1, \dots , \underline r_n),\ & M \to M[\underline{f(r_1,\dots,r_n)}/\alpha] \\
  \text{if } M \mid \alpha = \tY \lambda x. N,\ & M \to M[\lambda z. N[(\tY \lambda x. N)/x] z/\alpha]\\ \text{where }z&\text{ is not free in $N$}\\
  \text{if } M \mid \alpha = \tif{\underline r < 0}{N_1}{N_2},\ & M \to M[N_1/\alpha] \text{ where }r < 0 \\
  \text{if } M \mid \alpha = \tif{\underline r < 0}{N_1}{N_2},\ & M \to M[N_2/\alpha] \text{ where }r \geq 0 \\
  \text{if } M \mid \alpha = \tsample \text{ and $\lambda$ does}&\text{ not occur after $@_2$ or $\tY$ in $\alpha$},\\ & M \to M[\underline r/\alpha] \text{ where } r \in [0,1].
\end{align*}
In each of these cases, $M \mid \alpha$ is the \emph{redex}, and the reduction takes place at $\alpha$.
\end{definition}

@@

The reflexive transitive closure $\sim^*$ of this relation is used to define the set of \emph{potential positions} $L(M) = L_0(M) / \sim^*$, and each equivalence class can be considered as the same position as it may occur across multiple reachable skeletons. 
If $(N,\alpha) \sim^* (O,\beta)$, then $N \mid \alpha$ and $O \mid \beta$ both have the same shape (i.e.~they're either both the placeholder $\skeletonPlaceholder$, both variables, both applications, both $\tsample$s etc.), therefore it's well-defined to talk of the set of potential positions where there is a $\tsample$, $L_s(M)$. 
Formally $L_s(M) := \{[(X, \alpha)]_{\sim^\ast_M} : X \mid \alpha = \tsample\}$.
The new sample space is then defined as $I^{L_s(M)}$, with the Borel $\sigma$-algebra and product measure.
Since $I^{L_s(M)}$ is a countable product, the measure space is well-defined \cite[Cor.~2.7.3]{AshDD00}.

@@

\medskip
The reduction relation $\Rightarrow$ is nondeterministic, so it admits multiple possible reduction strategies. 
A \emph{reduction strategy} starting from a closed term $M$ is a measurable partial function $f$ from $\mathit{Rch}(M)$ to positions, such that for any reachable term $N$ where $f$ is defined, $f(N)$ is a position of a redex in $N$, and if $f(N)$ is not defined, $N$ is a value.
Using a reduction strategy $f$, a subset of $\Rightarrow$ that isn't nondeterministic, $\Rightarrow_f$, can be defined by $(N,s) \Rightarrow_f (N',s')$ just if $(N,s) \Rightarrow (N',s')$ and $N$ reduces to $N'$ with the redex at $f(N)$.

@@

More explicitly, the modified forms of the theorems are:
\begin{definition}
An \emph{antitone ranking function on $M$ with respect to a reduction strategy $r$} is a measurable function $f:\mathit{Rch}_r(M) \to \mathbb{R}$ such that $f(N) \geq 0$ for all $N$, and there exists an antitone function $\epsilon : \nnReal \to \pReal$ such that
\begin{itemize}
    \item $f(N) \geq \epsilon(f(N)) + f(N')$ if $N[r(N)]$ is a $\tY$-redex and $N$ reduces to $N'$ at $r(N)$
    \item $f(N) \geq \int_I f(N[\underline{x}/r(N)]) \, \Leb(\mathrm{d}x)$ if $N[r(N)] = \tsample$

    \item $f(N) \geq f(N')$ if $r(N)$ is any other redex, where $N \to N'$ at $r(N)$.
\end{itemize}
Any closed term for which an antitone ranking function with respect to a reduction strategy $r$ exists is called \emph{antitone rankable} with respect to $r$. 
\end{definition}

\begin{definition}
An \emph{antitone sparse ranking function on $M$ with respect to a reduction strategy $r$} is a partial function $f : \mathit{Rch}_r(M) \rightharpoonup \mathbb R$ such that there exists some antitone function $\epsilon : \nnReal \to \pReal$ such that
\begin{inparaenum}[(i)]
    \item $f(N) \geq 0$ for all $N \in \dom(f)$,
    \item $M \in \dom(f)$,
    \item for any $N \in \dom(f)$, evaluation of $N$ at the positions specified by $r$ will eventually reach some $O$ such that either $r(O)$ isn't defined or $f(O)$ is, and $f(N) \geq \mathbb E[f(O) + \epsilon(f(O)) \times \text{the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of $\dom(f)$).
\end{inparaenum}
\iffalse
\begin{itemize}
    \item $f(N) \geq 0$ for all $N$ where $f$ is defined.
    \item $f$ is defined at $M$.
    \item For any $N$ in the domain of definition of $f$, evaluation of $N$ at the positions specified by $r$ will eventually reach some $O$ such that either $r(O)$ isn't defined or $f(O)$ is, and $f(N) \geq \mathbb E[f(O) + \epsilon(f(O)) \times \text{the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
\end{itemize}
\fi
\end{definition}

If the antitone function is just the constant function $1$ instead, then it is called a \emph{(sparse) ranking function on $M$ with respect to $r$} instead, and the term is called \emph{rankable} with respect to $r$.

>>>>

A \defn{position} is a finite sequence of steps into a term, defined inductively as
\begin{align*}
\alpha ::= & \; {\cdot} \mid \lambda ; \alpha \mid @_1 ; \alpha \mid @_2 ; \alpha \mid \underline f_i ; \alpha \\
& {} \mid \tY ; \alpha \mid \textsf{if}_1 ; \alpha \mid \textsf{if}_2 ; \alpha \mid \textsf{if}_3 ; \alpha.
\end{align*}
The \defn{subterm of $M$ at $\alpha$}, denoted $M \mid \alpha$, is defined as
\begin{align*}
M \mid \cdot & = M \\
\lambda x. M \mid \lambda ; \alpha & = M \mid \alpha \\
M_1 M_2 \mid @_i ; \alpha & = M_i \mid \alpha \quad \text{for } i = 1,2 \\
\underline f(M_1,\dots,M_n) \mid \underline f_i ; \alpha & = M_i \mid \alpha \quad \text{for }i \leq n \\
\tY M \mid \tY ; \alpha & = M \mid \alpha \\
\tif{M_1 < 0}{M_2}{M_3} \mid \textsf{if}_i ; \alpha & = M_i \mid \alpha \quad \text{for } i = 1,2,3
\end{align*}
so that every subterm is located at a unique position, but not every position corresponds to a subterm (e.g. $x \, y \mid \lambda$ is undefined). 
A position such that $M\mid \alpha$ does exist is said to \emph{occur} in $M$. 
%\changed[lo]{
\emph{Substitution} of $N$ at position $\alpha$ in $M$, written $M[N/\alpha]$, is defined similarly.
For example, let 
\(
M = \lambda x \, y. y \, (\tif{x < 0}{y \, (\underline f (x))}{\underline 3})
\)
and
\(
\alpha =\lambda ; \lambda ; @_2 ; \textsf{if}_2 ; @_2
\)
then 
\(
M[\tsample / \alpha] = \lambda x \, y. y \, (\tif{x < 0}{y \, \tsample}{\underline 3}).
\)
%}

@@

With this notation, a more general reduction relation $\to$ can be defined.
\begin{definition}
\label{def:more general red}
The binary relation $\to$ is defined by the following rules, each is conditional on a redex occurring at position $\alpha$ in the term $M$:
\begin{align*}
  \text{if } M \mid \alpha = (\lambda x.N) V,\ & M \to M[N[V/x]/\alpha] \\
  \text{if } M \mid \alpha = \underline f (\underline r_1, \dots , \underline r_n),\ & M \to M[\underline{f(r_1,\dots,r_n)}/\alpha] \\
  \text{if } M \mid \alpha = \tY \lambda x. N,\ & M \to M[\lambda z. N[(\tY \lambda x. N)/x] z/\alpha]\\ \text{where }z&\text{ is not free in $N$}\\
  \text{if } M \mid \alpha = \tif{\underline r < 0}{N_1}{N_2},\ & M \to M[N_1/\alpha] \text{ where }r < 0 \\
  \text{if } M \mid \alpha = \tif{\underline r < 0}{N_1}{N_2},\ & M \to M[N_2/\alpha] \text{ where }r \geq 0 \\
  \text{if } M \mid \alpha = \tsample \text{ and $\lambda$ does}&\text{ not occur after $@_2$ or $\tY$ in $\alpha$},\\ & M \to M[\underline r/\alpha] \text{ where } r \in [0,1].
\end{align*}
In each of these cases, $M \mid \alpha$ is the \defn{redex}, and the reduction takes place at $\alpha$.
\end{definition}

@@

The reflexive transitive closure $\sim^*$ of this relation is used to define the set of \defn{potential positions} $L(M) = L_0(M) / \sim^*$, and each equivalence class can be considered as the same position as it may occur across multiple reachable skeletons. 
If $(N,\alpha) \sim^* (O,\beta)$, then $N \mid \alpha$ and $O \mid \beta$ both have the same shape (i.e.~they're either both the placeholder $\skeletonPlaceholder$, both variables, both applications, both $\tsample$s etc.), therefore it's well-defined to talk of the set of potential positions where there is a $\tsample$, $L_s(M)$. 
Formally $L_s(M) := \{[(X, \alpha)]_{\sim^\ast_M} : X \mid \alpha = \tsample\}$.
The new sample space is then defined as $I^{L_s(M)}$, with the Borel $\sigma$-algebra and product measure.
Since $I^{L_s(M)}$ is a countable product, the measure space is well-defined \cite[Cor.~2.7.3]{AshDD00}.

@@

\medskip
The reduction relation $\Rightarrow$ is nondeterministic, so it admits multiple possible reduction strategies. 
%\begin{definition}
A \defn{reduction strategy} starting from a closed term $M$ is a measurable partial function $f$ from $\mathit{Rch}(M)$ to positions, such that for any reachable term $N$ where $f$ is defined, $f(N)$ is a position of a redex in $N$, and if $f(N)$ is not defined, $N$ is a value.
Using a reduction strategy $f$, a subset of $\Rightarrow$ that isn't nondeterministic, $\Rightarrow_f$, can be defined by $(N,s) \Rightarrow_f (N',s')$ just if $(N,s) \Rightarrow (N',s')$ and $N$ reduces to $N'$ with the redex at $f(N)$.
%\end{definition}

@@

More explicitly, the modified forms of the theorems are:
\begin{definition}
An \defn{antitone ranking function on $M$ with respect to a reduction strategy $r$} is a measurable function $f:\mathit{Rch}_r(M) \to \mathbb{R}$ such that $f(N) \geq 0$ for all $N$, and there exists an antitone function $\epsilon : \nnReal \to \pReal$ such that
\begin{itemize}
    \item $f(N) \geq \epsilon(f(N)) + f(N')$ if $N[r(N)]$ is a $\tY$-redex and $N$ reduces to $N'$ at $r(N)$
    \item $f(N) \geq \int_I f(N[\underline{x}/r(N)]) \, \Leb(\mathrm{d}x)$ if $N[r(N)] = \tsample$

    \item $f(N) \geq f(N')$ if $r(N)$ is any other redex, where $N \to N'$ at $r(N)$.
\end{itemize}
Any closed term for which an antitone ranking function with respect to a reduction strategy $r$ exists is called \defn{antitone rankable} with respect to $r$. 
\end{definition}

\begin{definition}
An \defn{antitone sparse ranking function on $M$ with respect to a reduction strategy $r$} is a partial function $f : \mathit{Rch}_r(M) \rightharpoonup \mathbb R$ such that there exists some antitone function $\epsilon : \nnReal \to \pReal$ such that
\begin{inparaenum}[(i)]
    \item $f(N) \geq 0$ for all $N \in \dom(f)$,
    \item $M \in \dom(f)$,
    \item for any $N \in \dom(f)$, evaluation of $N$ at the positions specified by $r$ will eventually reach some $O$ such that either $r(O)$ isn't defined or $f(O)$ is, and $f(N) \geq \mathbb E[f(O) + \epsilon(f(O)) \times \text{the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of $\dom(f)$).
\end{inparaenum}
\iffalse
\begin{itemize}
    \item $f(N) \geq 0$ for all $N$ where $f$ is defined.
    \item $f$ is defined at $M$.
    \item For any $N$ in the domain of definition of $f$, evaluation of $N$ at the positions specified by $r$ will eventually reach some $O$ such that either $r(O)$ isn't defined or $f(O)$ is, and $f(N) \geq \mathbb E[f(O) + \epsilon(f(O)) \times \text{the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
\end{itemize}
\fi
\end{definition}

If the antitone function is just the constant function $1$ instead, then it is called a \defn{(sparse) ranking function on $M$ with respect to $r$} instead, and the term is called \defn{rankable} with respect to $r$.
